A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
conclusion

Truly understanding how programmers use a programming language is critical to building usable and effective language features.
When working on a general purpose languages (such as JavaScript), we as programmers can appeal to our own intuitions, as many of us have had at least some interaction with said languages.
This intuition does not extend to domain-specific languages (DSLs), which are not really designed for members of our community.
As more and more domain-experts, who may not have a traditional programming background, are exposed to the practice of programming, DSLs are developed to suit their needs.

The R programming language is an example of this.
R was developed to foster an easy and interactive programming experience, and the language is not very restrictive and highly expressive.
Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
Further, R is very slow, and the poor performance quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.

Types are a perfect way to both make a language easier to debug and make it run faster.
For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
If we're interested in developing an easily adoptable type system for R language, we must ensure that it captures how existing code is being used.
That is precisely what we have accomplished in this work.

Throughout this paper, we have built up an understanding of types and polymorphism in R by analyzing the usage of types across a large corpus of R code.
With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
We analyzed the tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.

>>>>

Truly understanding how programmers use a programming language is critical
to building usable and effective language features.  When working on a
general purpose languages (such as JavaScript), we as programmers can appeal
to our own intuitions, as many of us have had at least some interaction with
said languages.  This intuition does not extend to domain-specific languages
(DSLs), which are not really designed for members of our community.

The R programming language is an example of this.  R was developed to foster
an easy and interactive programming experience. For this reason R is very
flexible and expressive.  Unfortunately, these liberties come at a cost:
debugging R code is challenging. Further, performance quickly becomes
prohibitive if programmers are not careful.

Types can make a language easier to debug and run faster.  For debugging,
some level of static type enforcement would give programmers more direct
feedback.  As for performance, type information enables optimizations. If we
are interested in developing an easily adoptable type system for R, we must
ensure that it captures how existing code is being used.

Throughout this paper, we have built up an understanding of types and
polymorphism in R by analyzing the usage of types across a large corpus of
code.  We highlighted some patterns of type usage, and developed a number of
type systems with differing levels of granularity. We analyzed the tests,
examples, and vignettes of over ten thousand packages on CRAN (over
\PERCENTCRAN of available packages), and traced over 15 million lines of R
code.
